---
title: "FIA + Climate Analysis Notebook"
output: html_notebook
---

Libraries:

```{r, warning = F, message = F}
library(tidyverse)
library(caret)
library(xgboost)
```


Read in the data:

```{r}
forest.df <- read.csv("ForestClimate.csv", header = TRUE)
str(forest.df)
```

Recall:

Looking at the top 10 trees by quantity, we have:

  1) 746 is quaking aspen,
  2) 93 Engelmann spruce,
  3) 108 lodgepole pine,
  4) 814 Gambel oak,
  5) 19 subalpine fir


Remove some of the identifier codes, turn species code into one-hot columns and re-scale the continuous variables:

```{r}
fst.df <- 
forest.df %>%
  
  ## remove extra columns
  select(-c(X, PLT_CN, 
            SOIL_PLT_CN, OVEN_DRY_SOIL_WT, TOTAL_WATER_CONTENT_PCT,
            C_ORG_PCT, C_INORG_PCT, TXTRLYR1, TXTRLYR2, 
            CONDID, 
            SPCD_202, SPCD_122, SPCD_65, SPCD_106, SPCD_18)) %>%
  
  ## rename species:
  rename(SPCD_aspen = SPCD_746,
         SPCD_spruce = SPCD_93,
         SPCD_pine = SPCD_108,
         SPCD_oak = SPCD_814,
         SPCD_fir = SPCD_19) %>%
  
   # create dummy variables for the top 5 species
   mutate(SPCD_aspen  = ifelse(SPCD_aspen  >= 1, 1, 0),
          SPCD_spruce = ifelse(SPCD_spruce >= 1, 1, 0),
          SPCD_pine   = ifelse(SPCD_pine   >= 1, 1, 0),
          SPCD_oak    = ifelse(SPCD_oak    >= 1, 1, 0),
          SPCD_fir    = ifelse(SPCD_fir    >= 1, 1, 0)) %>%
  
  # create elevation strata
  mutate(ELEV = round(ELEV, -2)) %>%

  ## re-scale continuous variables
  mutate_at(vars(ELEV,FIELD_MOIST_SOIL_WT:PPTW), ~scale(.)) %>%
  
  ## drop na's
  na.omit()

head(fst.df)
```

Import and re-scale the prediction grid data by the mean and standard deviation of the original data:

```{r}
pred.grid <- read.csv("prediction_grid.csv", header = TRUE) %>% 
  select(-c(X, SOIL_PLT_CN,OVEN_DRY_SOIL_WT, TOTAL_WATER_CONTENT_PCT,
            C_ORG_PCT, C_INORG_PCT, TXTRLYR1, TXTRLYR2, TXTRLYR1, TXTRLYR2, CONDID))

pred.grid <- pred.grid %>%
  mutate( 
         ELEV = (ELEV - mean(forest.df$ELEV))/sd(forest.df$ELEV),
         
         ## SOIL VARIABLES
         FIELD_MOIST_SOIL_WT = (FIELD_MOIST_SOIL_WT - mean(forest.df$FIELD_MOIST_SOIL_WT)) / sd(forest.df$FIELD_MOIST_SOIL_WT),
         
         BULK_DENSITY = (BULK_DENSITY - mean(forest.df$BULK_DENSITY)) / sd(forest.df$BULK_DENSITY),
         
         C_TOTAL_PCT = (C_TOTAL_PCT - mean(forest.df$C_TOTAL_PCT)) / sd(forest.df$C_TOTAL_PCT),
         N_TOTAL_PCT = (N_TOTAL_PCT - mean(forest.df$N_TOTAL_PCT)) / sd(forest.df$N_TOTAL_PCT),
         
         FORFLTHK = (FORFLTHK - mean(forest.df$FORFLTHK)) / sd(forest.df$FORFLTHK),
         LTRLRTHK = (LTRLRTHK - mean(forest.df$LTRLRTHK)) / sd(forest.df$LTRLRTHK),
         DPTHSBSL = (DPTHSBSL - mean(forest.df$DPTHSBSL)) / sd(forest.df$DPTHSBSL),
         
         ## CLIMATE VARIBLES
         AVGT = (AVGT - mean(forest.df$AVGT)) / sd(forest.df$AVGT),
         JANT = (JANT - mean(forest.df$JANT)) / sd(forest.df$JANT),
         JULT = (JULT - mean(forest.df$JULT)) / sd(forest.df$JULT),
         PPT  = (PPT  - mean(forest.df$PPT )) / sd(forest.df$PPT),
         VARS = (VARS - mean(forest.df$VARS)) / sd(forest.df$VARS),
         VARW = (VARW - mean(forest.df$VARW)) / sd(forest.df$VARW),
         VARA = (VARA - mean(forest.df$VARA)) / sd(forest.df$VARA),
         DIFS = (DIFS - mean(forest.df$DIFS)) / sd(forest.df$DIFS),
         DIFW = (DIFW - mean(forest.df$DIFW)) / sd(forest.df$DIFW),
         PPTS = (PPTS - mean(forest.df$PPTS)) / sd(forest.df$PPTS),
         PPTW = (PPTW - mean(forest.df$PPTW)) / sd(forest.df$PPTW))

head(pred.grid)        
```

Create data matrices for boosting:

```{r}
## split fst.df into a training and testing set
set.seed(321)
num.obs <- dim(fst.df)[1]
split.pct <- 0.8
trn.index <- sample(1:num.obs, round(split.pct*num.obs))

fst.train <- as.data.frame(fst.df[ trn.index,])
fst.test  <- as.data.frame(fst.df[-trn.index,])

## create xgb.Dmatrix objects for testing and training each of the species:

## ASPENS

XGaspen.trn <- xgb.DMatrix(data  = as.matrix(fst.train %>% 
                                               select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.train$SPCD_aspen))

XGaspen.tst <- xgb.DMatrix(data  = as.matrix(fst.test %>% 
                                               select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.test$SPCD_aspen))

XGaspen.full <- xgb.DMatrix(data = as.matrix(fst.df %>% select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.df$SPCD_aspen))


## ENGELMANN SPRUCE

XGspruce.trn <- xgb.DMatrix(data  = as.matrix(fst.train %>% 
                                                select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.train$SPCD_spruce))

XGspruce.tst <- xgb.DMatrix(data  = as.matrix(fst.test %>% 
                                                select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.test$SPCD_spruce))

XGspruce.full <- xgb.DMatrix(data = as.matrix(fst.df %>% 
                                                select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label  = as.matrix(fst.df$SPCD_spruce))


## LODGEPOLE PINE

XGpine.trn <- xgb.DMatrix(data  = as.matrix(fst.train %>% 
                                              select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.train$SPCD_pine))

XGpine.tst <- xgb.DMatrix(data  = as.matrix(fst.test %>% 
                                              select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.test$SPCD_pine))

XGpine.full <- xgb.DMatrix(data  = as.matrix(fst.df %>% 
                                               select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label  = as.matrix(fst.df$SPCD_pine))

## GAMBEL OAK

XGoak.trn <- xgb.DMatrix(data  = as.matrix(fst.train %>% 
                                             select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.train$SPCD_oak))
XGoak.tst <- xgb.DMatrix(data  = as.matrix(fst.test %>% 
                                             select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.test$SPCD_oak))

XGoak.full <- xgb.DMatrix(data  = as.matrix(fst.df %>% 
                                              select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label  = as.matrix(fst.df$SPCD_oak))


## DOUGLAS FIR

XGfir.trn <- xgb.DMatrix(data  = as.matrix(fst.train %>% 
                                             select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.train$SPCD_fir))

XGfir.tst <- xgb.DMatrix(data  = as.matrix(fst.test %>% 
                                             select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label = as.matrix(fst.test$SPCD_fir))

XGfir.full <- xgb.DMatrix(data  = as.matrix(fst.df %>% 
                                              select(ELEV,FIELD_MOIST_SOIL_WT:PPTW)),
                           label  = as.matrix(fst.df$SPCD_fir))


## GRID for PREDICTIONS

XG.predGrid <- xgb.DMatrix(data = as.matrix(pred.grid %>% select(-c(LON,LAT))))

```


Helper functions to determine the optimum depth and learning rate of XGBoosting for a fixed number of rounds using the precision-recall AUC metric.

```{r}
xg.optimize <- function(XG.trn, depth.seq, eta.seq, nround = 10, cv.folds = 10, resp.type = "binary:logistic"){
  
  ## create search grid
  search.grid <- cbind(expand.grid(depth = depth.seq, 
                                   eta = eta.seq),
                       train.auc = 1:(length(depth.seq)*length(eta.seq)),
                       test.auc = 1:(length(depth.seq)*length(eta.seq)))
  
  ## begin iterations
  for (i in 1:dim(search.grid)[1]){
    
    ## use xg cross-validation function 
    t <- xgb.cv(data = XG.trn,
                
                ## Paramaters to be optimized
                max.depth = search.grid$depth[i], 
                eta = search.grid$eta[i],
                
                ## folds and rounds
                nfold = cv.folds,
                nrounds = nround,
                
                ## control over fitting
                max_delta_step = 2,
                scale_pos_weight = 1.5,
                subsample = 0.8,
                
                ## objective to optimize
                stratified = TRUE,
                metrics = "auc",
                objective = resp.type,
              
                ## quiet output
                verbose = 0)
    
    ## put auc value in the search grid
    result.train <- t$evaluation_log$train_auc_mean
    result.test <- t$evaluation_log$test_auc_mean

    search.grid$train.auc[i] <- result.train[length(result.train)]
    search.grid$test.auc[i] <- result.test[length(result.test)]
    
    print(paste("Iteration #: ", i))
    
    }
  
  ## return search values
  return(search.grid)

}

# xg.tune_rounds <- function(XG.trn, XG.val, depth, eta, nround, early.stop, cv.folds = 10) {
#    t <- xgb.cv(data = XG.trn,
#                 max.depth = depth, 
#                 eta = eta,
#                 nfold = cv.folds,
#                 nrounds = nround,
#                 subsample = 0.75,
#                 ## control over fitting
#                 max_delta_step = 2,
#                 scale_pos_weight = 0.8,
#                 metrics = "aucpr",
#                 objective = "binary:logistic",
#                 
#                 ## watches to stop 
#                 watchlist = list(train = XG.trn, val = XG.val),
#                 early_stopping_rounds = early.stop)
#    
#    return(t)
#   
# }
# 
# ## get best value
# 
# xg.param <- function(df){
#     
#   auc.test.max <- max(df$test.auc)
#   best.index <- which(df$test.auc == auc.test.max)
#   train.auc <- df$train.auc[best.index]
#   best.depth <- df$depth[best.index]
#   best.eta <- df$eta[best.index]
#   
#   return.df <- data.frame(auc.test = auc.test.max,
#                           train.auc = train.auc,
#                           depth = best.depth,
#                           eta = best.eta)
#   
#   ## return hyper-parameters and auc
#   return(return.df)
#   
# }

```


## Optimize hyper-parameters for the tree species

Get best hyper parameters for each of the tree species using a (non-exhaustive) grid search:

```{r, eval = F}
set.seed(43)
depth.vals <- c(3,4,6,8,10)
eta.vals   <- c(0.1,0.3,0.5,1) #10^seq(-2, 1.5, length = 6)

aspen.optim <- xg.optimize(XG.trn = XGaspen.trn,
                           cv.folds = 10,
                           depth.seq = depth.vals,
                           eta.seq = eta.vals)

spruce.optim <- xg.optimize(XG.trn = XGspruce.trn,
                            cv.folds = 10,
                            depth.seq = depth.vals,
                            eta.seq = eta.vals)

pine.optim <- xg.optimize(XG.trn = XGpine.trn,
                          cv.folds = 10,
                          depth.seq = depth.vals,
                          eta.seq = eta.vals)

oak.optim <- xg.optimize(XG.trn = XGoak.trn,
                         cv.folds = 10,
                         depth.seq = depth.vals,
                         eta.seq = eta.vals)

fir.optim <- xg.optimize(XG.trn = XGfir.trn,
                         cv.folds = 10,
                         depth.seq = depth.vals,
                         eta.seq = eta.vals)

aspen.optim
spruce.optim
pine.optim
oak.optim
fir.optim
```
* Aspen: depth = 4, eta = 0.3
* Spruce : depth = 3, eta = 0.3
* Pine: depth = 3, eta = 0.5
* Oak: depth = 3, eath = 0.3
* Fir: depth = 3, eta = 0.3


Generally, the tendency is to over-fit the data.



## Examining Aspen Predictions

Model fitting and prediction accuracy:

```{r}
xg.aspen <- xgboost(data = XGaspen.trn,
                    max.depth = 4, 
                    eta = 0.3,
                    nrounds = 10,
                    
                    ## control over fitting
                    max_delta_step = 2,
                    scale_pos_weight = 1.5,
                    
                    objective = "binary:logistic")

xg.yhat.trn <- predict(xg.aspen, newdata = XGaspen.trn, type = "response")
xg.yhat.trn <- as.numeric(xg.yhat.trn >= 0.5)

xg.yhat <- predict(xg.aspen, newdata = XGaspen.tst, type = "response")
xg.yhat <- as.numeric(xg.yhat >= 0.5)

mean(xg.yhat == fst.test$SPCD_aspen)
mean(xg.yhat.trn == fst.train$SPCD_aspen)

MLmetrics::F1_Score(factor(xg.yhat), factor(fst.test$SPCD_aspen))
MLmetrics::F1_Score(factor(xg.yhat.trn), factor(fst.train$SPCD_aspen))
```

Make a spatial/map plot:

```{r}
## Prediction on full data
xgF.aspen <- xgboost(data = XGaspen.full,
                    max.depth = 4, 
                    eta = 0.3,
                    nrounds = 10,
                    
                    ## control over fitting
                    max_delta_step = 2,
                    scale_pos_weight = 0.8,
                    
                    objective = "binary:logistic")

## Make predictions on the prediction grid
aspen.preds <- predict(xgF.aspen, newdata = XG.predGrid, type = response)

aspen.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(aspen.preds = ifelse(aspen.preds > 0.5, 1, 0))


aspen.pts <-  fst.df %>%
  select(LON, LAT, SPCD_aspen) %>%
  mutate(SPCD_aspen = factor(SPCD_aspen))


aspen.plot <- 
ggplot(data = aspen.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = aspen.preds), interpolate = TRUE) +
  scale_fill_gradient(low= "white", high="darkgreen", guide = 'none') +
  geom_point(data = aspen.pts, aes(x = LON, y = LAT, color = SPCD_aspen),size = .4, alpha = 1) +
  scale_color_discrete(name = "Observed", labels = c("Absent", "Present")) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1))+
  ggtitle('Predicted Quaking Aspen Distribution')
  
aspen.plot
```


Get a list of variable importance:

```{r}
xgb.importance(model = xg.aspen)
```

## Examining Spruce Predictions:


Fit a model and make predictions:



Model fitting and prediction accuracy:

```{r}
xg.spruce <- xgboost(data = XGspruce.trn,
                     max.depth = 3, 
                     eta = 0.3,
                     nrounds = 10,
                     ## control over fitting
                     max_delta_step = 2,
                     scale_pos_weight = 1.5,
                     objective = "binary:logistic")

xg.yhat.trn <- predict(xg.spruce, newdata = XGspruce.trn, type = "response")
xg.yhat.trn <- as.numeric(xg.yhat.trn > 0.35)

xg.yhat <- predict(xg.spruce, newdata = XGspruce.tst, type = "response")
xg.yhat <- as.numeric(xg.yhat > 0.35)

mean(xg.yhat == fst.test$SPCD_spruce)
mean(xg.yhat.trn == fst.train$SPCD_spruce)

MLmetrics::F1_Score(factor(xg.yhat), factor(fst.test$SPCD_spruce))
MLmetrics::F1_Score(factor(xg.yhat.trn), factor(fst.train$SPCD_spruce))
```

Plot predictions:

```{r}
xg.spruce <- xgboost(data = XGspruce.full,
                    max.depth = 3, 
                    eta = 0.3,
                    nrounds = 10,
                    
                    ## control over fitting
                    max_delta_step = 2,
                    scale_pos_weight = 1.5,
                    
                    objective = "binary:logistic")

## Make predictions on the prediction grid
spruce.preds <- predict(xg.spruce, newdata = XG.predGrid, type = response)

spruce.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(spruce.preds = ifelse(spruce.preds>0.35,1,0))


spruce.pts <-  fst.df %>%
  select(LON, LAT, SPCD_spruce) %>%
  mutate(SPCD_spruce = factor(SPCD_spruce))

spruce.plot <- 
ggplot(data = spruce.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = spruce.preds), interpolate = TRUE) +
  scale_fill_gradient(low= "white", high="darkgreen", guide = 'none') +
  geom_point(data = spruce.pts, aes(x = LON, y = LAT, color = SPCD_spruce),size = .4, alpha = 1) +
  scale_color_discrete(name = "Observed", labels = c("Absent", "Present")) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) +
  ggtitle("Predicted Engelmann Spruce Distribution")

spruce.plot
```

Get a list of variable importance:

```{r}
xgb.importance(model = xg.spruce)
```

## Examining Pine


Model fitting and prediction accuracy:

```{r}
xg.pine <- xgboost(data = XGpine.trn,
                    max.depth = 3, 
                    eta = 0.5,
                    nrounds = 10,
                    ## control over fitting
                    max_delta_step = 2,
                    scale_pos_weight = 1.5,
                    objective = "binary:logistic")

xg.yhat.trn <- predict(xg.pine, newdata = XGpine.trn, type = "response")
xg.yhat.trn <- as.numeric(xg.yhat.trn > 0.25)

xg.yhat <- predict(xg.pine, newdata = XGpine.tst, type = "response")
xg.yhat <- as.numeric(xg.yhat > 0.25)

mean(xg.yhat == fst.test$SPCD_pine)
mean(xg.yhat.trn == fst.train$SPCD_pine)

MLmetrics::F1_Score(factor(xg.yhat), factor(fst.test$SPCD_pine))
MLmetrics::F1_Score(factor(xg.yhat.trn), factor(fst.train$SPCD_pine))
```

Plot predictions:

```{r}
xg.pine <- xgboost(data = XGpine.full,
                    max.depth = 3, 
                    eta = 0.5,
                    nrounds =  10,
                    
                    ## control over fitting
                    max_delta_step = 2,
                    scale_pos_weight = 1.5,
                    
                    objective = "binary:logistic")

## Make predictions on the prediction grid
pine.preds <- predict(xg.pine, newdata = XG.predGrid, type = response)

pine.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(pine.preds = ifelse(pine.preds>0.25, 1, 0))


pine.pts <-  fst.df %>%
  select(LON, LAT, SPCD_pine) %>%
  mutate(SPCD_pine = factor(SPCD_pine))

pine.plot <- 
ggplot(data = pine.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = pine.preds), interpolate = TRUE) +
  scale_fill_gradient(low = 'white', high="darkgreen", guide = 'none') +
  geom_point(data = pine.pts, aes(x = LON, y = LAT, color = SPCD_pine),size = .4, alpha = 1) +
  scale_color_discrete(name = "Observed", labels = c("Absent", "Present")) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) +
  ggtitle('Predicted Lodgepole Pine Distribution')

pine.plot
```

Get a list of variable importance:

```{r}
xgb.importance(model = xg.pine)
```

## Examining Oak

Model fitting and prediction accuracy:

```{r}
xg.oak <- xgboost(data = XGoak.trn,
                    max.depth = 5, 
                    eta = 0.3,
                    nrounds = 8,
                    ## control over fitting
                    max_delta_step = 2,
                    scale_pos_weight = 1.5,
                    objective = "binary:logistic")

xg.yhat.trn <- predict(xg.oak, newdata = XGoak.trn, type = "response")
xg.yhat.trn <- as.numeric(xg.yhat.trn > 0.35)

xg.yhat <- predict(xg.oak, newdata = XGoak.tst, type = "response")
xg.yhat <- as.numeric(xg.yhat > 0.35)

mean(xg.yhat == fst.test$SPCD_oak)
mean(xg.yhat.trn == fst.train$SPCD_oak)

MLmetrics::F1_Score(factor(xg.yhat), factor(fst.test$SPCD_oak))
MLmetrics::F1_Score(factor(xg.yhat.trn), factor(fst.train$SPCD_oak))
```

Plot predictions:

```{r}
xg.oak <- xgboost(data = XGoak.full,
                    max.depth = 3, 
                    eta = 0.3,
                    nrounds = 10,
                    
                    ## control over fitting
                    max_delta_step = 2,
                    scale_pos_weight = 1.5,
                    
                    objective = "binary:logistic")

## Make predictions on the prediction grid
oak.preds <- predict(xg.oak, newdata = XG.predGrid, type = response)

oak.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(oak.preds = ifelse(oak.preds>0.35,1,0))


oak.pts <-  fst.df %>%
  select(LON, LAT, SPCD_oak) %>%
  mutate(SPCD_oak = factor(SPCD_oak))

oak.plot <- 
ggplot(data = oak.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = oak.preds), interpolate = TRUE) +
  scale_fill_gradient(low= "white", high="darkgreen") +
  geom_point(data = oak.pts, aes(x = LON, y = LAT, color = SPCD_oak),size = .2, alpha = 0.2)+
  scale_color_discrete(name = "Observed", labels = c("Absent", "Present")) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) +
  ggtitle('Predicted Gambel Oak Distribution')

oak.plot
```

Get a list of variable importance:

```{r}
xgb.importance(model = xg.oak)
```

## Examining FIR

Model fitting and prediction accuracy:

```{r}
xg.fir <- xgboost(data = XGfir.tst,
                    max.depth = 3, 
                    eta = 0.3,
                    nrounds = 10,
                    ## control over fitting
                    max_delta_step = 3,
                    scale_pos_weight = 1.5,
                    objective = "binary:logistic")

xg.yhat.trn <- predict(xg.fir, newdata = XGfir.trn, type = "response")
xg.yhat.trn <- as.numeric(xg.yhat.trn > 0.35)

xg.yhat <- predict(xg.fir, newdata = XGfir.tst, type = "response")
xg.yhat <- as.numeric(xg.yhat > 0.35)

mean(xg.yhat == fst.test$SPCD_fir)
mean(xg.yhat.trn == fst.train$SPCD_fir)

MLmetrics::F1_Score(factor(xg.yhat), factor(fst.test$SPCD_fir))
MLmetrics::F1_Score(factor(xg.yhat.trn), factor(fst.train$SPCD_fir))
```

Plot predictions:

```{r}
xg.fir <- xgboost(data = XGfir.full,
                    max.depth = 3, 
                    eta = 0.3,
                    nrounds = 10,
                    
                    ## control over fitting
                    max_delta_step = 3,
                    scale_pos_weight = 1.5,
                    
                    objective = "binary:logistic")

## Make predictions on the prediction grid
fir.preds <- predict(xg.fir, newdata = XG.predGrid, type = response)

fir.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(fir.preds = ifelse(fir.preds>0.35, 1, 0))


fir.pts <-  fst.df %>%
  select(LON, LAT, SPCD_fir) %>%
  mutate(SPCD_fir = factor(SPCD_fir))

fir.plot <- 
ggplot(data = fir.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = fir.preds), interpolate = TRUE) +
  scale_fill_gradient(low= "white", high="darkgreen", guide = 'none') +
  geom_point(data = fir.pts, aes(x = LON, y = LAT, color = SPCD_fir),size = .4, alpha = 1) +
  scale_color_discrete(name = "Observed", labels = c("Absent", "Present")) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) +
  ggtitle('Predicted Subalpine Fir Distribution')


fir.plot
```

Get a list of variable importance:

```{r}
xgb.importance(model = xg.fir)
```

Make a nice looking plot of spruce, pine and fir distribution plots:

```{r, fig.width=12}
ggpubr::ggarrange(aspen.plot, spruce.plot, pine.plot, fir.plot, 
                  ncol = 2, nrow = 2, 
                  common.legend = TRUE,
                  legend = 'bottom')
```



## Modeling distribution given a change in DIFS

```{r}
## sequence of different difference in summer temperatures
difs.offset <- seq(from = mean(pred.grid$DIFS) - 2*sd(fst.df$DIFS),
                   to = mean(pred.grid$DIFS) + 2*sd(fst.df$DIFS),
                   length.out = 100)

pine.ratio <- 1:length(difs.offset)
difs.pred <- pred.grid$DIFS #make a copy of original prediction

## current total number of squares filled as denominator
denom.pine <- sum(pine.pred.grid$pine.preds)


for (i in 1:length(difs.offset)){
  pine.preds.difs <- 0
  ## make xg prediction grid and add offset
  XG.gridAlt <- xgb.DMatrix(data = as.matrix(pred.grid %>%
                                               select(-c(LON,LAT)) %>%
                                               mutate(DIFS = difs.pred + difs.offset[i])))
  
  pine.preds.difs <- ifelse(predict(xg.pine, newdata = XG.gridAlt, type = response) > 0.25, 1,0)
  
  pred.grid.tmp <- pred.grid %>% mutate(pine.preds.difs = 
                                          ifelse(predict(xg.pine, newdata = 
                                                           XG.gridAlt, type = response) > 0.25, 
                                                 1,0)) %>%
    filter(ELEV >= (7000 - mean(forest.df$ELEV))/sd(forest.df$ELEV))
  
  pine.ratio[i] = sum(pred.grid.tmp$pine.preds.difs) / denom.pine
  
}

pine.change <- data.frame(DDIFS = difs.offset, DIST_RATIO = pine.ratio)

ggplot(data = pine.change, aes(x = DDIFS, y = DIST_RATIO)) + geom_point() + geom_line()
```

Calculate the actual temperature difference:

```{r}
sd(forest.df$DIFS) 
```


Distribution for 1-SD decrease in DIFS:

```{r}
XG.gridAlt <- xgb.DMatrix(data = as.matrix(pred.grid %>%
                                               select(-c(LON,LAT)) %>%
                                               mutate(DIFS = difs.pred - 1)))

## Make predictions on the prediction grid
pine.preds <- predict(xg.pine, newdata = XG.gridAlt, type = response)

pine.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(pine.preds = ifelse(pine.preds>0.25, 1, 0))


pine.pts <-  fst.df %>%
  select(LON, LAT, SPCD_pine) %>%
  mutate(SPCD_pine = factor(SPCD_pine))

pine.plot.dec <- 
ggplot(data = pine.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = pine.preds), interpolate = TRUE) +
  scale_fill_gradient(low = 'white', high="darkgreen", guide = 'none') +
  geom_point(data = pine.pts, aes(x = LON, y = LAT, color = SPCD_pine),size = .4, alpha = 1) +
  scale_color_discrete(name = "Observed", labels = c("Absent", "Present")) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) +
  ggtitle('Decrease in DIFS Pine Distribution')

pine.plot.dec
```

Distribution for 1-SD increase in DIFS:

```{r}
XG.gridAlt <- xgb.DMatrix(data = as.matrix(pred.grid %>%
                                               select(-c(LON,LAT)) %>%
                                               mutate(DIFS = difs.pred + 1 )))

## Make predictions on the prediction grid
pine.preds <- predict(xg.pine, newdata = XG.gridAlt, type = response)

pine.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(pine.preds = ifelse(pine.preds>0.5, 1, 0))


pine.pts <-  fst.df %>%
  select(LON, LAT, SPCD_pine) %>%
  mutate(SPCD_pine = factor(SPCD_pine))

pine.plot.inc <- 
ggplot(data = pine.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = pine.preds), interpolate = TRUE) +
  scale_fill_gradient(low = 'white', high="darkgreen", guide = 'none') +
  geom_point(data = pine.pts, aes(x = LON, y = LAT, color = SPCD_pine),size = .4, alpha = 1) +
  scale_color_discrete(name = "Observed", labels = c("Absent", "Present")) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) +
  ggtitle('Increase in DIFS Pine Distribution')

pine.plot.inc
```

Final plot:

```{r, fig.width=12}
ggpubr::ggarrange(pine.plot.dec, pine.plot.inc,
                  ncol = 2, nrow = 1, 
                  common.legend = TRUE,
                  legend = 'right')
```



Let's do something similar for fir by manipulating VARA:

Decrease in VARA:

```{r}
vara.pred <- pred.grid$VARA #make a copy of original prediction

XG.gridAlt <- xgb.DMatrix(data = as.matrix(pred.grid %>%
                                               select(-c(LON,LAT)) %>%
                                               mutate(VARA = vara.pred - 1)))

## Make predictions on the prediction grid
fir.preds <- predict(xg.fir, newdata = XG.gridAlt, type = response)

fir.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(fir.preds = ifelse(fir.preds>0.35, 1, 0))


fir.pts <-  fst.df %>%
  select(LON, LAT, SPCD_fir) %>%
  mutate(SPCD_fir = factor(SPCD_fir))

fir.plot.dec <- 
ggplot(data = fir.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = fir.preds), interpolate = TRUE) +
  scale_fill_gradient(low = 'white', high="darkgreen", guide = 'none') +
  geom_point(data = fir.pts, aes(x = LON, y = LAT, color = SPCD_fir),size = .4, alpha = 1) +
  scale_color_discrete(name = "Observed", labels = c("Absent", "Present")) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) +
  ggtitle('Decrease in VARA fir Distribution')

fir.plot.dec

```

Increase in VAR for fir:

```{r}
XG.gridAlt <- xgb.DMatrix(data = as.matrix(pred.grid %>%
                                               select(-c(LON,LAT)) %>%
                                               mutate(VARA = vara.pred + 1)))

## Make predictions on the prediction grid
fir.preds <- predict(xg.fir, newdata = XG.gridAlt, type = response)

fir.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(fir.preds = ifelse(fir.preds>0.35, 1, 0))


fir.pts <-  fst.df %>%
  select(LON, LAT, SPCD_fir) %>%
  mutate(SPCD_fir = factor(SPCD_fir))

fir.plot.inc <- 
ggplot(data = fir.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = fir.preds), interpolate = TRUE) +
  scale_fill_gradient(low = 'white', high="darkgreen", guide = 'none') +
  geom_point(data = fir.pts, aes(x = LON, y = LAT, color = SPCD_fir),size = .4, alpha = 1) +
  scale_color_discrete(name = "Observed", labels = c("Absent", "Present")) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) +
  ggtitle('Increase in VARA fir Distribution')

fir.plot.inc

```

Final plot:

```{r, fig.width=12}
ggpubr::ggarrange(fir.plot.dec, fir.plot.inc,
                  ncol = 2, nrow = 1, 
                  common.legend = TRUE,
                  legend = 'right')
```


## Pine & Fir Abundance with a Tweedie Response

Create a new data-frame and XG matrices:


```{r}
## Create new data frame with species counts and relavent variables

abundance.df <- fst.df %>%
  select(-c(LAT, LON, SPCD_aspen:SPCD_pine)) %>%
  mutate(SPCD_pine = forest.df$SPCD_108,
         SPCD_fir = forest.df$SPCD_19)

## Split into test and training
set.seed(231)
num.obsA <- dim(abundance.df)[1]
split.pctA <- 0.8
trn.indexA <- sample(1:num.obs, round(split.pctA*num.obsA))

abundance.train <- as.data.frame(abundance.df[ trn.indexA,])
abundance.test  <- as.data.frame(abundance.df[-trn.indexA,])

## Pine boosting matrices

## LODGEPOLE PINE

XGApine.trn <- xgb.DMatrix(data  = as.matrix(abundance.train %>% 
                                              select(ELEV:PPTW)),
                           label = as.matrix(abundance.train$SPCD_pine))

XGApine.tst <- xgb.DMatrix(data  = as.matrix(abundance.test %>% 
                                              select(ELEV:PPTW)),
                           label = as.matrix(abundance.test$SPCD_pine))

XGApine.full <- xgb.DMatrix(data  = as.matrix(abundance.df %>% 
                                               select(ELEV:PPTW)),
                           label  = as.matrix(abundance.df$SPCD_pine))

## Subalpine fir

XGAfir.trn <- xgb.DMatrix(data  = as.matrix(abundance.train %>% 
                                              select(ELEV:PPTW)),
                           label = as.matrix(abundance.train$SPCD_fir))

XGAfir.tst <- xgb.DMatrix(data  = as.matrix(abundance.test %>% 
                                              select(ELEV:PPTW)),
                           label = as.matrix(abundance.test$SPCD_fir))

XGAfir.full <- xgb.DMatrix(data  = as.matrix(abundance.df %>% 
                                               select(ELEV:PPTW)),
                           label  = as.matrix(abundance.df$SPCD_fir))
```

Count number of zero responses:

```{r}
abundance.df %>% select(SPCD_pine) %>% filter(SPCD_pine > 0) %>% count() / dim(abundance.df)[1]
```



Modify the cross-validation function for Tweedie regression:

```{r}
xg.optimizeT <- function(XG.trn, depth.seq, eta.seq, nround = 4, cv.folds = 10){
  
  ## create search grid
  search.grid <- cbind(expand.grid(depth = depth.seq, 
                                   eta = eta.seq),
                       train.loglik = 1:(length(depth.seq)*length(eta.seq)),
                       test.loglik = 1:(length(depth.seq)*length(eta.seq)))
  
  ## begin iterations
  for (i in 1:dim(search.grid)[1]){
    
    ## use xg cross-validation function 
    t <- xgb.cv(data = XG.trn,
                
                ## Paramaters to be optimized
                max.depth = search.grid$depth[i], 
                eta = search.grid$eta[i],
                
                ## folds and rounds
                nfold = cv.folds,
                nrounds = nround,
                
                ## control over fitting
                max_delta_step = 2,
                scale_pos_weight = 1.5,
                sub_sample = 0.8,
                
                ## objective to optimize
                stratified = TRUE,
                metrics = "tweedie-nloglik@1.5",
                objective = "reg:tweedie",
              
                ## quiet output
                verbose = 0)
    
    ## put auc value in the search grid
    result.train <- t$evaluation_log$`train_tweedie_nloglik@1.5_mean`
    result.test <- t$evaluation_log$`test_tweedie_nloglik@1.5_mean`

    search.grid$train.loglik[i] <- result.train[length(result.train)]
    search.grid$test.loglik[i] <- result.test[length(result.test)]
    
    print(paste("Iteration #: ", i))
    
    }
  
  ## return search values
  return(search.grid)

}
```



## Pine Abundance

```{r, eval = F}
a.depth <- c(2,3,4,6,8,10)
a.eta <- c(0.1,0.3,0.5,1)

pine.optim <- xg.optimizeT(XG.trn = XGApine.trn,
                          cv.folds = 10,
                          nround = 10,
                          depth.seq = a.depth,
                          eta.seq = a.eta)

pine.optim
```


Model fitting and prediction accuracy:

```{r}
xgA.pine <- xgboost(data = XGApine.trn,
                    max.depth = 3, 
                    eta = 0.1,
                    nrounds = 10,
                    ## control over fitting
                    max_delta_step = 2,
                    tweedie_variance_power = 1.5,
                    objective = "reg:tweedie")


xg.yhat.trn <- predict(xgA.pine, newdata = XGApine.trn, type = "response")
xg.yhat <- predict(xg.pine, newdata = XGApine.tst, type = "response")

mean((abundance.train$SPCD_pine - xg.yhat.trn)^2)
mean((abundance.test$SPCD_pine - xg.yhat)^2)
```

```{r}
xgA.pine <- xgboost(data = XGApine.full,
                    max.depth = 3, 
                    eta = 0.3,
                    nrounds = 10,
                    ## control over fitting
                    max_delta_step = 2,
                    tweedie_variance_power = 1.5,
                    objective = "reg:tweedie")

## Make predictions on the prediction grid
pineA.preds <- predict(xgA.pine, newdata = XG.predGrid, type = response)

pineA.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(pine.preds = pineA.preds)


pine.pts <-  fst.df %>%
  select(LON, LAT, SPCD_pine) %>%
  mutate(SPCD_pine = factor(SPCD_pine))

pine.abund.plot <- 
ggplot(data = pineA.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = pine.preds), interpolate = TRUE) +
  scale_fill_gradient(low= "white", high="darkgreen", name = "Pine Abundance") +
  geom_point(data = pine.pts, aes(x = LON, y = LAT, color = SPCD_pine),size = .2, alpha = 0.3) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) + 
  scale_color_discrete(name = "Observed Pine", labels = c("Absent", "Present")) +
  ggtitle("Predicted Lodgepole Pine Abundance") +
  guides(fill = guide_colorbar(order = 1), color = guide_legend(order= 0))
 
pine.abund.plot 
```

```{r}
xgb.importance(model = xgA.pine)
```

## Fir Abundance

```{r, eval = F}
a.depth <- c(3,4,6,8,10)
a.eta <- c(0.1,0.3,0.5,1)

fir.optim <- xg.optimizeT(XG.trn = XGAfir.trn,
                          cv.folds = 10,
                          nround = 10,
                          depth.seq = a.depth,
                          eta.seq = a.eta)

fir.optim
```



Model fitting and prediction accuracy:

```{r}
xgA.fir <- xgboost(data = XGAfir.trn,
                    max.depth = 3, 
                    eta = 0.3,
                    nrounds = 10,
                    ## control over fitting
                    max_delta_step = 2,
                    tweedie_variance_power = 1.5,
                    objective = "reg:tweedie")


xg.yhat.trn <- predict(xgA.fir, newdata = XGAfir.trn, type = "response")
xg.yhat <- predict(xg.fir, newdata = XGAfir.tst, type = "response")

mean((abundance.train$SPCD_fir - xg.yhat.trn)^2)
mean((abundance.test$SPCD_fir - xg.yhat)^2)
```


Prediction and map:

```{r}
xgA.fir <- xgboost(data = XGAfir.full,
                    max.depth = 3, 
                    eta = 0.3,
                    nrounds = 10,
                    ## control over fitting
                    max_delta_step = 2,
                    tweedie_variance_power = 1.5,
                    objective = "reg:tweedie")

## Make predictions on the prediction grid
firA.preds <- predict(xgA.fir, newdata = XG.predGrid, type = response)

firA.pred.grid <- pred.grid %>% select(LAT, LON) %>% mutate(fir.preds = firA.preds)


fir.pts <-  fst.df %>%
  select(LON, LAT, SPCD_fir) %>%
  mutate(SPCD_fir = factor(SPCD_fir))

fir.abund.plot <- 
ggplot(data = firA.pred.grid, aes(x = LON, y = LAT)) + 
  geom_raster(aes(fill = fir.preds), interpolate = TRUE) +
  scale_fill_gradient(low= "white", high="darkgreen", name = 'Fir Abundance') +
  geom_point(data = fir.pts, aes(x = LON, y = LAT, color = SPCD_fir),size = .2, alpha = 0.3) +
  borders("county","colorado",colour="grey70") + 
  borders("state", size = 2) +
  scale_x_continuous(breaks = seq(from = -109.5, to = 101.25, by = 1)) +
  coord_fixed(ratio = 1.25, xlim = c(-108.95,-102.2), ylim = c(37,41.1)) +
  scale_color_discrete(name = "Observed Fir", labels = c("Absent", "Present")) +
  ggtitle("Predicted Subalpine Fir Abundance") +
  guides(fill = guide_colorbar(order = 1), color = guide_legend(order= 0))


fir.abund.plot
```


```{r}
xgb.importance(model = xgA.fir)
```



Final plot:

```{r, fig.width=12}
ggpubr::ggarrange(pine.abund.plot, fir.abund.plot,
                  ncol = 2, nrow = 1)
```




