---
title: "Wrangling FIA and Climate Data"
output: html_notebook
---

Libraries:

```{r, warning = FALSE, message = FALSE}
library(rFIA)
library(dplyr)
library(ggplot2)
library(rpart)
library(mgcv)
library(tidyr)
library(RNetCDF)
library(tidync)
library(ncmeta)
library(dplyr)
library(ggplot2)
library(tools)
library(sf)
library(maps)
library(patchwork)
```



## Determine top 10 most abundant trees in CO:

Load in Forest data and join `TREE` and `PLOT` tables:

(he `TREE` table contains a good deal of the relevant information regarding the trees in a plot area. We also may want to look at other tables too, which may contain relevant information. In order to get the location (in geographic coordinates) of a plot, this table must be joined with `PLOT` table where the `PLT_CN` is a foreign key linking the tables together. As we can see, there are over 200 different variables in the `TREE` table.))

```{r}
## Object holding loaction of data
COforest <- readFIA("C:/Users/jddru/Desktop/ASAENVR/data/co-fia")

## Plot and Tree tables:
CO.TREE <- COforest$TREE
CO.PLOT <- COforest$PLOT

## Wrangle and Join Tables
CO.tree <- CO.TREE %>% select(
  ## Grab ID and location/plot information...
  CN, PLT_CN, INVYR, ##UNITCD, COUNTYCD, PLOT, SUBP, TREE, CONDID, AZIMUTH, DIST,
  ## Relavent tree measurment...
  STATUSCD, SPCD, DIA, DIAHTCD, HT, HTCD, CR, CCLCD)
  ## Enviromental factors
  ##AGENTCD, CULL, DECAYCD)

CO.plot <- CO.PLOT %>% select(
  ## Grab ID and location information..
  CN, LAT, LON, ELEV
)

## Join the subsets of tables together

CO.forest <- full_join(CO.tree, CO.plot, by = c("PLT_CN" = "CN"))
head(CO.forest)
```


Let's count the number of species of trees in Colorado and see which are the most prevalent:

```{r}
CO.forest %>% filter(INVYR>=2002 & STATUSCD == 1) %>% 
  count(SPCD, sort = TRUE) %>% 
  mutate(percent = round(100 * n / sum(n),3))
```


Looking at the top 10 trees by valsuantity, we have:

  1) 746 is valsuaking aspen (my favorite),
  2) 93 Engelmann spruce,
  3) 108 lodgepole pine,
  4) 814 Gambel oak,
  5) 19 subalpine fir
  6) 106 Common or two-needle pinion
  7) 65 Utah Juniper
  8) 202 Douglas-fir
  9) 122 Ponderosa Pine
  10) 18 Corkbark fir
  
Let's filter by these 10 species:

```{r}
spcd.id <- c(746,93,108,814,19,106,65,202,122,18)
CO.forestSub <- CO.forest %>% filter(SPCD %in% spcd.id)
head(CO.forestSub) 
```

Get plot ID for live trees (`STATUSCD` = 1) from 2002 and beyond:

```{r}
#CO.TreeBin <- CO.forestSub %>%
  # mutate(value = 1) %>%
  # pivot_wider(names_from = SPCD,
  #             values_from = value,
  #             values_fill = list(value = 0),
  #             names_prefix = "SPCD") %>%
  ##mutate(value = 1) %>% spread(key = SPCD, value, fill = 0) %>%
  #filter(INVYR >= 2002 & STATUSCD == 1) %>%
  #select(-c(CN))

CO.TreeBin <- CO.forestSub %>%
  filter(INVYR >= 2002 & STATUSCD == 1) %>%
  select(PLT_CN, INVYR, LAT, LON, ELEV, SPCD) %>%
  mutate(SPCD = factor(SPCD)) %>%
  group_by(PLT_CN, LAT, LON, ELEV, SPCD) %>%
  summarise(species_count =  n()) %>%
  pivot_wider(names_from = SPCD,
              values_from = species_count,
              values_fill = list(species_count = 0),
              names_prefix = 'SPCD_') %>%
  ungroup()



head(CO.TreeBin)
```



Plot locations with no observations:

```{r}
no.obs <- CO.forest %>% filter((CN %in% CO.TreeBin$PLT_CN) == FALSE &
                                 is.na(INVYR) == TRUE)
```


## Soils data

Read into R:
```{r}
soil.lab <- read.csv("C:/Users/jddru/Desktop/ASAENVR/data/co-fia/CO_SOILS_LAB.csv", header = TRUE)
soil.loc <- read.csv("C:/Users/jddru/Desktop/ASAENVR/data/co-fia/CO_SOILS_SAMPLE_LOC.csv", header = TRUE)

soil.lab.summary <- soil.lab %>%
  group_by(CN, PLT_CN, STATECD, COUNTYCD, PLOT) %>%
  summarise_at(vars(FIELD_MOIST_SOIL_WT:OLSEN_P), mean)


soil.geo <- inner_join(CO.TreeBin, soil.lab.summary)

ggplot(soil.geo, aes(x = LON, y = LAT)) +
  geom_point()

naniar::vis_miss(soil.lab.summary, warn_large_data = FALSE)
```
Remove the mineral measurements due the missing data.

```{r}
soil.geo.sub <- soil.geo %>%
  select(PLT_CN, LON, LAT, FIELD_MOIST_SOIL_WT, OVEN_DRY_SOIL_WT, TOTAL_WATER_CONTENT_PCT,
         BULK_DENSITY, C_ORG_PCT, C_INORG_PCT, C_TOTAL_PCT, N_TOTAL_PCT) %>%
  mutate(C_INORG_PCT = C_TOTAL_PCT-C_ORG_PCT) %>%
  na.omit() %>%
  group_by(PLT_CN, LON, LAT) %>%
  summarise_all(mean)


naniar::vis_miss(soil.geo.sub, warn_large_data = FALSE)
```

Get the texture measurements at the locations:
```{r}
## Select appropriate columns
soil.loc.sub <- soil.loc %>%
  filter(SOILS_STATCD == 1) %>%
  select(PLT_CN, FORFLTHK, LTRLRTHK, CONDID, TXTRLYR1, TXTRLYR2, DPTHSBSL) %>%
  mutate(DPTHSBSL = replace(DPTHSBSL, DPTHSBSL==99.9,0)) %>%
  na.omit() %>%
  group_by(PLT_CN,TXTRLYR1,TXTRLYR2) %>%
  summarise_all(mean) %>%
  mutate(TXTRLYR1 = factor(TXTRLYR1),
         TXTRLYR2 = factor(TXTRLYR2))
  

naniar::vis_miss(soil.loc.sub, warn_large_data = FALSE)
```


Join:

```{r}
soils.df <- inner_join(soil.geo.sub,soil.loc.sub) 

naniar::vis_miss(soils.df, warn_large_data = FALSE)
```

Look at the summary of the combined soils data:

```{r}
summary(soils.df)
```

Plot the soil locations in black, and the tree plot locations in red:

```{r}
ggplot(data = CO.TreeBin, aes(x = LON, y = LAT)) +
  geom_point(color = "red") +
  geom_point(data = soils.df, aes(x = LON, y = LAT))
```

Interpolate the soil data with KNN k=1 interpolation. Based on the latitude and longitude coordinates of a tree plot location, we will find the nearest soil plot and obtain the plot ID:

```{r}
tree.locs <- data.frame(CO.TreeBin %>% select(c(PLT_CN,LON,LAT)))
tree.locs <- tree.locs[,2:3]
soil.locs <- soils.df %>% select(c(PLT_CN,LON,LAT)) %>% as.data.frame()
soil.coords <- soil.locs[,2:3]
soil.cs <- factor(soil.locs[,1])

soil.interp <- class::knn(soil.coords, 
                          tree.locs,
                          soil.cs, 
                          k = 1)
```


Add soil plot column:

```{r}
CO.TreeBin$SOIL_PLT_CN = as.numeric(as.character(soil.interp))

head(CO.TreeBin)
```

Join soil data by soil plot ID:

```{r}
CO.TreeBinSoil <- inner_join(CO.TreeBin, soils.df, by = c("SOIL_PLT_CN" = "PLT_CN")) %>%
  select(-c(LON.y, LAT.y)) %>%
  rename(LAT = LAT.x, LON = LON.x)

head(CO.TreeBinSoil)
```

valsuick plot to check that things make sense:

```{r}
ggplot(data = CO.TreeBinSoil, aes(x = LON, y = LAT, color = C_ORG_PCT)) +
  geom_point()
```

## Climate Data

Set file paths do data:

```{r}
file.maxt <- "C:/Users/jddru/Desktop/ASAENVR/data/era-interim/MAXT.nc" # Maximum temperature
file.mint <- "C:/Users/jddru/Desktop/ASAENVR/data/era-interim/MINT.nc" # Minimum temperature
file.prec <- "C:/Users/jddru/Desktop/ASAENVR/data/era-interim/PREC.nc" # Precipitation

##source('C:/Users/jddru/Desktop/ASAENVR/Colorado_Forests/load_netcdf.R')
```

Functions to get the data from the netCDF file:

```{r}
## function to get the data

getCDF.data <- function(filePath, lonRange, latRange, timeRange){
  
  ## values for selecting a lat/long range and a time
  
  vals <- t(data.frame(lon = lonRange, 
                    lat = latRange, 
                    tim = timeRange, 
                    row.names = c("min", "max")))
  
  ## Load the file
  
  df.out <- tidync(filePath)
  
  ## Get subset based on lon/lat and time specifications, return what's below
  
  df.out %>% hyper_filter(lon = between(lon, vals["lon", "min"], vals["lon", "max"]), 
                          lat = between(lat, vals["lat", "min"], vals["lat", "max"]), 
                           time = between(time, vals["tim", "min"], vals["tim", "max"])) %>% 
    hyper_tibble()

}

## function to get the times

ParseNetCDFTime <- function(variableMetadata, variableData) {
  
  ## get the time units from metatdata
  time.unit <- nc_atts(variableMetadata, "time") %>% 
    unnest(cols = c(value)) %>% 
    filter(name == "units")
  
  ## get the time parts and return what's below
  time.parts <- utcal.nc(time.unit$value, variableData$time)
  ISOdatetime(time.parts[, "year"],
              time.parts[, "month"],
              time.parts[, "day"],
              time.parts[, "hour"],
              time.parts[, "minute"],
              time.parts[, "second"],
              tz = "UTC")
}
```


We need to get human-usable times for aggregation and filtering, this needs to be done in a slightly round-about-way to avoid memory issues. (Thus far, I haven't thought of a more elegant way of doing this with our current NETCDF infrastructure.) First let's grab a time series at a location in order to get a mapping from the date index to an actual date:

```{r}
## Grab series
era.ts <- getCDF.data(filePath = file.maxt, 
                       lonRange = c(255,255), 
                       latRange = c(40.5,40.5), 
                       timeRange = c(0,14245))

## Convert the time index to date and drop unnecessary cols
era.ts <- era.ts %>% 
  mutate(date = ParseNetCDFTime(file.maxt, era.ts)) %>% 
  select(-(MAXT)) %>% 
  select(-(lon)) %>% 
  select(-(lat))

## Parse the date to year, month, date for filtering/aggregation
era.ts <- era.ts %>% mutate(year = lubridate::year(date),
                            month = lubridate::month(date),
                            day = lubridate::day(date))
head(era.ts)
```


Now we have a mapping from the the time index to a date. In order to set the indies for the `getCDF.data` function, we'll need to find the time index corresponding to Jan 1 2002. Note that the climate data only goes up through the end of 2017, so we can drop the years 2018 and 2019 from the FIA table. We also need to set-up lat/lon bounds over Colorado, found via Google maps (with some padding):

```{r}
t.start <- as.numeric(era.ts %>% filter(year == 1979, month == 1, day == 1) %>% 
                        select(time))
t.end <- max(era.ts$time)

lat.max <- 45
lat.min <- 33

lon.max <- 262.5
lon.min <- 247.75
```

Now let's grab the subset of the ERA-Interim climate data:

```{r}
maxt.df <- getCDF.data(filePath = file.maxt, 
                       lonRange = c(lon.min,lon.max), 
                       latRange = c(lat.min,lat.max), 
                       timeRange = c(t.start,t.end))
mint.df <- getCDF.data(filePath = file.mint, 
                       lonRange = c(lon.min,lon.max), 
                       latRange = c(lat.min,lat.max), 
                       timeRange = c(t.start,t.end))
prec.df <- getCDF.data(filePath = file.prec, 
                       lonRange = c(lon.min,lon.max), 
                       latRange = c(lat.min,lat.max), 
                       timeRange = c(t.start,t.end))
```

Now, let's join the date table to each of these so we can aggregate/filter:

```{r}
maxt.df <- inner_join(maxt.df, era.ts, by = "time")
mint.df <- inner_join(mint.df, era.ts, by = "time")
prec.df <- inner_join(prec.df, era.ts, by = "time")

climate.df <- left_join(maxt.df, mint.df) %>% 
  left_join(., prec.df) %>% 
  mutate(date = ParseNetCDFTime(file.maxt, maxt.df)) %>% 
  mutate(year = lubridate::year(date),
         month = lubridate::month(date),
         day = lubridate::day(date)) %>%
  select(lon, lat, MAXT, MINT, PREC, year, month, day)

head(climate.df)
```


Let's obtain the following explanatory variables:

* AVGT Mean annual temperature
* JANT Mean January temperature
* JULT Mean July temperature
* PPT Annual precipitation
* VARH Variance of high temperature
* VARL Variance of low temperature
* VARP Variance of precipitation


Main climate variables:

```{r}
## Average temperature
temp.avg <- climate.df %>% 
  group_by(lon, lat) %>%
  summarise(AVGT = mean((MAXT + MINT)/ 2 ))

## Average January/February temperature
temp.janAvg <- climate.df %>%
  filter(month %in% c(1,2)) %>%
  group_by(lon, lat) %>%
  summarise(JANT = mean((MAXT + MINT)/2))

## Average July/August temperature  
temp.julyAvg <- climate.df %>%
  filter(month %in% c(7,8)) %>%
  group_by(lon, lat) %>%
  summarise(JULT = mean((MAXT + MINT)/2))

## Variation in mean summer temperature
temp.varS <- climate.df %>%
  filter(month %in% c(6,7,8)) %>%
  group_by(lon,lat) %>%
  summarise(VARS = var((MAXT + MINT)/2))

## Variation in winter mean temperature
temp.varW <- climate.df %>%
  filter(month %in% c(1,2,12)) %>%
  group_by(lon,lat) %>%
  summarise(VARW = var((MAXT + MINT)/2))

## Variation in annual mean temperature
temp.varA <- climate.df %>%
  group_by(lon,lat) %>%
  summarise(VARA = var((MAXT + MINT)/2))

## Difference in summer high and low temperature
temp.diffS <- climate.df %>%
  filter(month %in% c(6,7,8)) %>%
  group_by(lon,lat) %>%
  summarise(DIFS = mean(MAXT - MINT))

## Difference in winter high and low temperature
temp.diffW <- climate.df %>%
  filter(month %in% c(1,2,12)) %>%
  group_by(lon,lat) %>%
  summarise(DIFW = mean(MAXT - MINT))

## Annual cumulative precipitation
temp.precAvg <- climate.df %>%
  group_by(lon,lat) %>%
  summarise(PPT = sum(PREC))

## Summer cumulative precipitation
temp.precS <- climate.df %>%
  filter(month %in% c(6,7,8)) %>%
  group_by(lon,lat) %>%
  summarise(PPTS = sum(PREC))

## Winter cumulative precipitation
temp.precW <- climate.df %>%
  filter(month %in% c(1,2,12)) %>%
  group_by(lon,lat) %>%
  summarise(PPTW = sum(PREC))

## Annual precipitation variation
temp.precvar <- climate.df %>%
  group_by(lon,lat) %>%
  summarise(VARP = var(PREC))

climateSummary.df <- left_join(temp.avg,temp.janAvg) %>%
  left_join(.,temp.julyAvg) %>%
  left_join(., temp.precAvg) %>%
  left_join(., temp.varS) %>%
  left_join(., temp.varW) %>%
  left_join(., temp.varA) %>%
  left_join(., temp.diffS) %>%
  left_join(., temp.diffW) %>%
  left_join(., temp.varW) %>%
  left_join(., temp.precAvg) %>%
  left_join(., temp.precS) %>%
  left_join(., temp.precW) %>%
  left_join(., temp.precvar) 

remove(temp.avg, temp.janAvg, temp.julyAvg, temp.precAvg, temp.highvar, temp.lowvar, temp.precvar)

head(climateSummary.df)
```

Fit GAM models to smooth each of the variables for interpolation/prediction:

```{r}
gam.avgt <- gam(AVGT ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
gam.jant <- gam(JANT ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML',data = climateSummary.df)
gam.jult <- gam(JULT ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
gam.ppt  <- gam(PPT  ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
gam.vars <- gam(VARS ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
gam.varw <- gam(VARW ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
gam.vara <- gam(VARA ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
gam.difs <- gam(DIFS ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
gam.difw <- gam(DIFW ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
gam.ppts <- gam(PPTS ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
gam.pptw <- gam(PPTW ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML',data = climateSummary.df)
gam.varp <- gam(VARP ~ s(lon) + s(lat) + ti(lon, lat), method = 'REML', data = climateSummary.df)
```

Look at difference in summer temperature:

```{r, fig.height = 4, fig.width = 10}
par(mfrow = c(1,3))
plot.gam(gam.difs, select = 1, residuals = TRUE, xlab = 'Longitude')
plot.gam(gam.difs, select = 2, residuals = TRUE, xlab = 'Latitude')
plot.gam(gam.difs, select = 3, residuals = TRUE, xlab = 'Longitude', ylab = 'Latitude')

par(mfrow = c(2,2))
gam.check(gam.difs)
```
Look at variation in precipitation:

```{r, fig.height = 4, fig.width = 10}
par(mfrow = c(1,3))
plot(gam.vara, residuals = TRUE)

par(mfrow = c(2,2))
gam.check(gam.vara)
```


Predict climate variables at the plot locations and append to the soil+forest data-frame:

```{r}
CO.TreeBinSoilClimate <- CO.TreeBinSoil %>%
  mutate(AVGT = predict(gam.avgt,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         JANT = predict(gam.jant,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         JULT = predict(gam.jult,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         PPT  = predict(gam.ppt,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         VARS = predict(gam.vars,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         VARW = predict(gam.varw,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         VARP = predict(gam.varp,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         VARA = predict(gam.vara,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         DIFS = predict(gam.difs,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)), 
         DIFW = predict(gam.difw,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         PPTS = predict(gam.ppts,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         PPTW  = predict(gam.pptw,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         VARP  = predict(gam.varp,
                        newdata = data.frame(lon = 360 + CO.TreeBinSoil$LON,
                                             lat = CO.TreeBinSoil$LAT)),
         )


summary(CO.TreeBinSoilClimate)
```


Check temperature to make sure prediction was okay:

```{r}
ggplot(data = CO.TreeBinSoilClimate, aes(x = LON, y = LAT, color = scale(AVGT))) +
  geom_point()
```

Write to a a csv for later or new notebook

```{r}
#write.csv(CO.TreeBinSoilClimate, "ForestClimate.csv")
```



Create a table for making predictions:

We need to get elevation information, soil information, and climate information on a grid of longitude/latitude points. 

Construct grid of Lat/Lon coordinates

```{r}
CO.plot.nona <- CO.plot %>% select(c(LON, LAT, ELEV)) 
CO.plot.nona <- CO.plot.nona[-which(is.na(CO.plot$ELEV)),]

lon.list <- seq(from = min(CO.plot.nona$LON),
                to = max(CO.plot.nona$LON),
                by = 0.01)
lat.list <- seq(from = min(CO.plot.nona$LAT),
                to = max(CO.plot.nona$LAT),
                by = 0.01)
coord.grid <- expand.grid(LON = lon.list, LAT = lat.list)
paste("coord grid size: ", dim(coord.grid))
head(coord.grid)
```

Use knn for nearest neighbor interpolation of elevation:

```{r}
elev.interp <- class::knn(CO.plot.nona %>% select(c(LON, LAT)), 
                          coord.grid,
                          CO.plot.nona$ELEV, 
                          k = 1)
coord.grid$ELEV <- as.numeric(as.character(elev.interp))
head(coord.grid)
```

Check with a visual raster plot of the data:

```{r}
ggplot(data = coord.grid, aes(x = LON, y = LAT, fill = ELEV)) + 
  geom_raster(interpolate = TRUE)
```

Using knn to interpolate the soil data as was done before, lets do the same:

```{r}
## Create the model interpolation
soilGrid.interp <- class::knn(soil.coords,
                              coord.grid[,c(1,2)],
                              soil.cs)

## Get the corresponding plot number for each lat/lon
coord.grid <- coord.grid %>% 
  mutate(SOIL_PLT_CN = 
           as.numeric(as.character(soilGrid.interp)))

## Join soil table with coordinate table
coord.grid <- inner_join(coord.grid, soils.df, by = c("SOIL_PLT_CN" = "PLT_CN")) %>%
  select(-c(LON.y, LAT.y)) %>%
  rename(LON = LON.x, LAT = LAT.x)

head(coord.grid)
```

Check with a visual raster plot of the data:

```{r}
ggplot(data = coord.grid, aes(x = LON, y = LAT, fill = N_TOTAL_PCT)) + 
  geom_raster(interpolate = TRUE)+
  scale_fill_continuous(low = "blue", high = "red")
```


Using the existing GAM climate smooths, we predict at the grid of LAT/LON locations:

```{r}
coord.grid <- coord.grid %>%
  mutate(AVGT = predict(gam.avgt,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         JANT = predict(gam.jant,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         JULT = predict(gam.jult,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         PPT  = predict(gam.ppt,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         VARS = predict(gam.vars,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         VARW = predict(gam.varw,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         VARP = predict(gam.varp,
                       newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         VARA = predict(gam.vara,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         DIFS = predict(gam.difs,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)), 
         DIFW = predict(gam.difw,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         PPTS = predict(gam.ppts,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         PPTW  = predict(gam.pptw,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         VARP  = predict(gam.varp,
                        newdata = data.frame(lon = 360 + coord.grid$LON,
                                             lat = coord.grid$LAT)),
         )
summary(coord.grid)

```



```{r}
ggplot(data = coord.grid, aes(x = LON, y = LAT, fill = PPTS)) + 
  geom_raster(interpolate = TRUE)
```

Write the prediction grid to a csv file for later use:

```{r}
#write.csv(coord.grid, "prediction_grid.csv")
```










